\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-04-17 doc/incertae_sedis_latex2.tex
Start reorganizing sections 3 + 4.

%% LyX 2.3.0 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=3cm}
\setlength{\parskip}{6pt}
\setlength{\parindent}{0pt}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[authoryear]{natbib}

@@

\title{\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae
sedis}}\texttt{ taxa}}
\maketitle

@@

We thus require that one of the input trees is a taxonomy tree. The
taxonomy has a number of features not found in the input phylogenies.
While the input trees need not cover the entire taxon set, we require
the taxonomy to be comprehensive. The taxonomy also has labels for
internal nodes, whereas input phylogenies only have labels for leaf
nodes. We refer to labels of taxonomy nodes as ``taxon names'',
and require that taxon names and taxonomy nodes have a one-to-one
correspondence. The taxonomy may also contain out-degree 1 nodes.
These nodes correspond to monotypic taxa which contain a single child
of lower taxonomic rank.

@@

The Open Tree of Life project seeks to build a comprehensive supertree
covering all of life. The approach is to combine information in published
phylogenies with a comprehensive taxonomy that supplies taxon names
\citep{redelings2017supertree}. We describe an extension to this
supertree method so that we can resolve taxonomic uncertainty by using
published phylogenies to place \emph{incertae sedis} taxa. We describe
an updated semantics for conflict between phylogeny edges and taxonomy
edges and an updated semantics for assigning taxonomy names to nodes
in the synthetic tree.

Our approach to \emph{incertae sedis} taxa enables us to use published
phylogenies to update the taxonomy by extending taxonomic groups to
include \emph{incertae sedis} taxa placed within them. The ability
to update the taxonomy is important, since the placement of \emph{incertae
sedis} taxa creates artificial conflicts with a non-updated taxonomy.
Thus, the ability to correctly assess compatibility with the taxonomy
when the taxonomy contains \emph{incertae sedis taxa} allows us to
include thousands of new taxa that were previously filtered out to
avoid conflict. It also enables us to include extinct taxa, since
many of these taxa are \emph{incertae sedis}. This makes substantial
progress towards our goal of completely representation of taxa.

@@

After a synthesis tree is constructed, we attach taxon names to the
synthesis tree for each non-conflicting taxon. These names attach
at the MRCA of the include group. Then, for each taxon $A$ that is
found on the synth tree, we associate with the first taxon $B$ found
in a rootward walk on the synthesis tree. If $A$ is not a descendant
of $B$ on the taxonomy, then we say that $A$ is \emph{placed} within
$B$ by the synthesis tree. While this may occur if $A$ is \emph{incertae
sedis}, more complex scenarios are possible. We note that if method
does not handle \emph{incertae sedis} taxa, then this scenario could
not occur because any taxa $B$ that $A$ is placed within would be
considered incompatible with the synthesis tree and so their names
would not be applied.

In order to fully describe a placement of $A$ within $B$ we consider
the sequence of taxa encountered on a walk from $A$ or $B$ up to
their MRCA $M$ on the taxonomy tree. Let us denote the sequence of
taxa encountered from $A$ as $A_{1},\ldots,A_{n},M$ where $A_{1}=A$.
Let $B_{1},\ldots,B_{m},M$ be the sequence of taxa encountered from
$B$ where $B_{1}=B$. Then $A_{n}$ must be \emph{incertae sedis},
and $A_{2},\ldots,A_{n}$ must be incompatible with the synthesis
tree (i.e. broken taxa). We then have that $A$ is placed successively
within $B_{m},B_{m-1},\ldots,B_{2},B_{1}$, and $m$ is the depth
of the placement of $A$.%
\begin{comment}
One question is whether this approach is able to handle nested incertae
sedis taxa.

@@

Synthesis with \emph{incertae sedis} taxa has the potential to resolve
uncertain taxon placements using information from phylogenies. The
propinquity pipeline has always been able to do this kind of resolution.
However, without special consideration given to \emph{naming}, placing
a taxon $A$ within a taxon $B$ results in conflict with taxon $B$
in the taxonomy. In order to avoid a situation where input phylogenies
conflict with a large number of taxa when \emph{incertae sedis} taxa
are placed within them, we have filtered \emph{incertae sedis} taxa
from the taxonomy when constructing all prior synthesis trees.

@@

\subsection{Correctly handling placement}

Thus, we seek a synthesis method that can correctly place taxa within
containing taxa without breaking the containing taxa. We change the
semantics of names to imply, not the exclusion of all non-included
taxa, but only some non-included taxa, as described in section \ref{sec:Semantics-of-incertae}.
As a result of this change in semantics, placing an IS taxon $A$
within a sister taxon $B$ no longer results in conflict with $B$.
This allows us to retain the split for $B$ within the synthesis tree,
so that taxonomy-only children of $B$ are correctly grouped with
their siblings that are referenced by the input trees. We may then
retain the name for the no-longer-broken taxon. Finally, we are then
able to stop filtering \emph{incertae sedis} taxa, so that they appear
in the synthesis tree. Thus, the synthesis tree is able to represent
substantially more species, without suffering the loss of taxa and
the loss of structure.

@@

\subsubsection{Conflicting placement among input trees}

@@

\subsubsection{Example B}

However, incertae sedis taxa are not always tip nodes, but may themselves
contain other taxa. In such cases, it is possible for input trees
to conflict with the incertae sedis taxon itself. For example, consider
the following example
\begin{itemize}
\item $T_{1}=((w_{1},w_{2},z_{1}),y_{1})$
\item $T_{2}=((y_{1},y_{2},z_{2}),w_{1})$
\item Taxonomy tree $((w_{1},w_{2})w,(y_{1},y_{2})y,(z_{1},z_{2},z_{3})z)$
with $z$ marked \emph{incertae sedis}.
\begin{itemize}
\item splits $w_{1}w_{2}|\bullet y_{1}y_{2}$, $y_{1}y_{2}|\bullet w_{1}w_{2}$
and $z_{1}z_{2}|\bullet w_{1}w_{2}y_{1}y_{2}$
\end{itemize}
\end{itemize}
In this case, tree $T_{1}$ places $z_{1}$ within $w$, while $T_{2}$
places $z_{2}$ within $y$. Since different members of $z$ are placed,
neither placement for $z$ is rejected. Instead the taxon $z$ is
broken, the name $z$ disappears, and $z_{3}$ floats to the top level.

Furthermore, since taxon $z$ is a broken \emph{incertae sedis }taxon,
all of its children are effectively incertae sedis independently,
with the difference that they cannot be placed within each other.
Therefore, the names $w$ and $y$ are not lost, since the taxa placed
within them are \emph{incertae sedis} names. Note that $z_{3}$ maybe
therefore be placed in a third location, since it is also \emph{incertae
sedis}.

The situation here would be different if the monophyly of $z$ was
supported by an input phylogeny.

\subsubsection{Example C - Nested \emph{incertae sedis} taxa}

In addition to incertae sedis taxa containing other taxa, it is also
possible for incertae sedis taxa to contain nested \emph{incertae
sedis} taxa. \textbf{What issues might this raise?}

\section{Cases}

@@

\begin{figure}

\hfill{}\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/tree2-shaded}

@@

\hfill{}\subfloat[Synthesis Tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/synth2-shaded}

@@

\caption{Broken incertae sedis clade\label{fig:Broken-incertae-sedis}. Taxonomy
tree (b) contains \emph{incertae sedis }taxon $C$ that is broken
by input tree (a). Supertree construction results in a synthesis tree
(c) that places $C1$ and $C2$ separately within $A$.}

@@

@@

\caption{Placing one incertae sedis group within another. Here I is placed
within A, and then AI is placed within A.}

@@

It would be nice to allow running conflict against the cleaned taxonomy,
though. One way to do this would be to generated a ``placed taxonomy'',
with groups extended to include \emph{incertae sedis} taxa that have
been placed within them. This would not require any updating to the
conflict-analysis code in the annotation step.

@@

This probably is outside the scope of the paper, but if we could come
up with a modified conflict algorithm, that would be nice/useful,
and probably novel. It would also probably be slower...

>>>>

%% LyX 2.3.0 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{luainputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=3cm}
\setlength{\parskip}{6pt}
\setlength{\parindent}{0pt}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[authoryear]{natbib}

@@

\title{\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae
sedis}}\texttt{ taxa}}

\author{Benjamin D. Redelings and Mark T. Holder}
\maketitle

@@

Here we describe a supertree method that takes a taxonomy as one of
its inputs. The taxonomy has a number of features not found in the
input phylogenies. While the input trees need not cover the entire
taxon set, our method requires that the taxonomy to be comprehensive.
The taxonomy also has labels for internal nodes, whereas input phylogenies
only have labels for leaf nodes. We refer to labels of taxonomy nodes
as ``taxon names'', and require that taxon names and taxonomy nodes
have a one-to-one correspondence. The taxonomy may also contain out-degree
1 nodes. These nodes correspond to monotypic taxa which contain a
single child of lower taxonomic rank.

@@

The Open Tree of Life project seeks to build a comprehensive supertree
covering all of life. The approach is to combine information in published
phylogenies with a comprehensive taxonomy that supplies taxon names
\citep{redelings2017supertree}. In practice, the method has been
used with the Open Tree of Life taxonomy \citep{rees2017automated}.We
describe an extension to this supertree method so that we can resolve
taxonomic uncertainty by using published phylogenies to place \emph{incertae
sedis} taxa. We describe an updated semantics for conflict between
phylogeny edges and taxonomy edges and an updated semantics for assigning
taxonomy names to nodes in the synthetic tree.

Our approach to \emph{incertae sedis} taxa enables us to use published
phylogenies to update the taxonomy by extending taxonomic groups to
include \emph{incertae sedis} taxa placed within them. The ability
to update the taxonomy is important, since the placement of \emph{incertae
sedis} taxa creates artificial conflicts with a non-updated taxonomy.
Thus, the ability to correctly assess compatibility with the taxonomy
when the taxonomy contains \emph{incertae sedis }taxa allows us to
include thousands of new taxa that were previously filtered out to
avoid conflict. It also enables us to include extinct taxa, since
many of these taxa are \emph{incertae sedis}. This makes substantial
progress towards our goal of completely representation of taxa.

@@

After we have attached taxon names to the synthesis tree, we would
like to interpret the position of these names in terms of placing
\emph{incertae sedis }taxa in some spot, possibly a new spot. This
would allow us to interpret the synthesis tree as saying that phylogenetic
information has (for example) placed genus A within family B, or perhaps
in a monotypic, unnamed family that contains only $A$. The simplest
approach to placement involves noting whenever a taxon $B$ is a descendant
of a taxon $A$ on the named synthesis tree but not the taxonomy tree.
For example, in Figure \ref{fig:Handling-incertae-sedis}, taxon $C$
is a descendant of taxon $A$ in the synthetic tree (Fig. \ref{fig:Handling-incertae-sedis}d),
but not the taxonomy (Fig. \ref{fig:Handling-incertae-sedis}b). Thus,
we could say that the synthetic tree places $C$ within $A$. Furthermore,
the synthetic tree additionally places $C$ within $AB$.

A slightly more general case is shown in Figure \ref{fig:Broken-incertae-sedis}.
Here taxon $C1$ is an descendant of $A$ in the synthesis tree, but
not the taxonomy. Thus $C1$ is placed within $A$. We note that the
most recent common ancestor (MRCA) of $A$ and $C1$ in the taxonomy
is the root node, and the path to the MRCA from $A$ is $A\to AB\to root$,
while the path to the MRCA from $C1$ is $C1\to C\to root$. Here
the additional node $AB$ on the path to the MRCA from $A$ indicates
that $C1$ is placed within $AB$ as well as $A$. In contrast, the
additional node $C$ on the path to the MRCA indicates that $C1$
is a broken \emph{incertae sedis} taxon. Thus we may consider an algorithm
that, for each taxon $B$ on the synthesis tree, finds the closest
ancestral taxon $A$, and checks if $A$ is an ancestor of $B$ on
the taxonomy. If not, we compute the paths $A\to A_{1}\to\ldots\to MRCA(A,B)$
and $B\to B_{1}\to\ldots\to MRCA(A,B)$.

However, this assumes that \emph{incertae sedis} taxa are always placed
within their siblings, or descendants of their siblings. As such,
it does not handle cases such as Figure \ref{fig:Placing-one-incertae},
in which $I$ is first placed as sibling to $AI$, and then $AI$
is placed within $I$. Note that $I$ is not a sibling of $AI$ or
a descendant of a sibling. Furthermore, the MRCA of $I$ and $AI$
is the root, and the path from $AI$ to the root includes $A$, but
$A$ is not broken. We suggest that this conundrum may be resolved
by first processing the placement of $I$ within $A$ by modifying
the taxonomy to reflect this placement\footnote{Does this actually work? Some placements break taxa.}.
When we consider the placement of $AI$ within $I$ on this modified
taxonomy, the MRCA of $I$ and $AI$ is now $A$ instead of the root,
and the correct interpretation results. We suggest that by walking
the tree in a pre-order fashion and processing shallowing placements
first, the previous approach of finding the MRCA and labelling intermediate
nodes as either additional placements or broken incertae sedis (or
incertae\_sedis\_inherited) taxa holds. Such an approach should also
handle cases containing nested incertae sedis taxa, as in Figure \ref{fig:Placing-one-incertae}.%
\begin{comment}
One question is whether this approach is able to handle nested incertae
sedis taxa.

@@

Synthesis with \emph{incertae sedis} taxa has the potential to resolve
uncertain taxon placements using information from phylogenies. Since
phylogenies are ranked higher than taxonomy, they may place members
of incertae sedis clades as they wish even when we do not take into
account the semantics of incertae sedis taxa. However, without the
incertae sedis semantics described above, placing a taxon $A$ within
a taxon $B$ results in conflict with taxon $B$ in the taxonomy.
This has three main effects:
\begin{enumerate}
\item The edge leading to clade $B$ is lost.
\item Any taxonomy-only taxa are placed at the MRCA of the phylogenetically
informed taxa, instead of at node $B$.
\item The name for taxon $B$ is lost.
\item We do not record $A$ as being placed within $B$.
\end{enumerate}
In order to avoid a situation where input phylogenies conflict with
a large number of taxa when \emph{incertae sedis} taxa are placed
within them, we have previously filtered \emph{incertae sedis} taxa
from the taxonomy when constructing all prior synthesis trees.

@@

We therefore seek a synthesis method that can correctly introduce
\emph{incertae sedis} taxa into containing taxa without breaking the
containing taxa. We change the semantics of names to imply, not the
exclusion of all non-included taxa, but only some non-included taxa,
as described in section \ref{sec:Semantics-of-incertae}. As a result
of this change in semantics, placing an IS taxon $A$ within a sister
taxon $B$ no longer results in conflict with $B$. This allows us
to retain the split for $B$ within the synthesis tree, so that taxonomy-only
children of $B$ are correctly grouped with their siblings that are
referenced by the input trees. We may then retain the name for the
no-longer-broken taxon. Finally, we are then able to stop filtering
\emph{incertae sedis} taxa, so that they appear in the synthesis tree.
Thus, the synthesis tree is able to represent substantially more species,
without suffering the loss of taxa and the loss of structure.

@@

\subsection{Conflicting placement among input trees}

{[}\textbf{BDR:} \emph{maybe we should merge this with Figure. \ref{fig:Broken-incertae-sedis}.
Although the point of figure \ref{fig:placement-example1} is to talk
about sub-problem decomposition as well as what the result looks like.}{]}

@@

@@

\begin{figure}
\hfill{}\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/tree1-shaded}

}\hfill{}\subfloat[Input tree \#2]{\includegraphics[scale=0.5]{Figures/is-clade/tree3-shaded}

@@

\hfill{}\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/is-clade/tax-shaded}

}\hfill{}\subfloat[Synthesis Tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/synth2-shaded}

@@

\caption{\label{fig:Broken-incertae-sedis} \emph{Incertae sedis} clade $C$
broken by conflicting placement. Input trees (a) and (b) conflict
in the placement of taxa in $C$. The synthesis supertree (d) places
places $C1$ and $C2$ separately within $A$, while unplaced taxa
float upwards. (\textbf{Note:} This is what the sub-problem solver
does. The entire pipeline would attach $C3$ at $A$ actually, since
it is the MRCA of $A1$ and $A2$. Should we change this?) }

@@

When the input phylogenies conflict on where members of an incertae
sedis clade should be placed, neither placement is rejected. Instead,
the taxon $C$ is broken, the name $C$ disappears, and $C3$ floats
to the top level. Algorithmic, alternatively placements of children
of an incertae sedis clade lead to contesting of all edges along with
taxa might be placed, creating a large, merged sub-problem, as in
Figure \ref{fig:placement-example1}.

In Figure \ref{fig:Broken-incertae-sedis}, tree $T_{1}$ places $C1$
next to $A1$, while $T_{2}$ places $C2$ next to $A2$. Since different
members of $C$ are placed, neither placement for $C$ is rejected.
Instead the taxon $C$ is broken, the name $C$ disappears, and $C3$
floats to the top level. Furthermore, since taxon $C$ is a broken
\emph{incertae sedis }taxon, all of its children are effectively incertae
sedis independently, with the exception that they cannot be placed
within each other. Therefore, the names $A$ and $AB$ are not lost,
since the taxa placed within them are \emph{incertae sedis} names.

The situation here would be different if the monophyly of $C$ was
supported by an high-ranked input phylogeny. In that case, instead
of breaking $C$, we would choose the placement in the highest-ranked
tree.

@@

\caption{\label{fig:Placing-one-incertae}Placing one incertae sedis group
within another. Here I is placed within A, and then AI is placed within
A.}

@@

It would be nice to allow running conflict against the cleaned taxonomy,
though. One way to do this would be to generated a ``placed taxonomy'',
with groups extended to include \emph{incertae sedis} taxa that have
been placed within them. This would not require any updating to the
conflict-analysis code in the annotation step.%
\begin{comment}

@@

This probably is outside the scope of the paper, but if we could come
up with a modified conflict algorithm, that would be nice/useful,
and probably novel. It would also probably be slower...
\end{comment}
